The ``Dark Side'' of Phylogenetic Comparative Methods

### 2014-08-24 Manuscript/dark.tex
Minor edits

A long time ago in a galaxy far, far away - also known as the 1980s - phylogenetic comparative methods (PCMs) were developed to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}). 
Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). 
PCMs have become extremely popular over recent years; the number of papers featured the phrase `phylogenetic comparative' has increased exponentially since the 1980s (Figure \ref{PCMCitations}). 
With new methods being published almost weekly, there has never been a better time to be a comparative biologist.

\begin{figure}[h]
\centering
\includegraphics[width = 8cm]{PCMCitations.png}
\caption{PCM papers through time}
\label{PCMCitations}
\end{figure}

Unfortunately, PCMs also have a ``dark side''; they make various assumptions and suffer from biases in exactly the same way as any statistical method \citep{freckleton2009seven,boettiger2012your}. 
The more popular these methods become, the less awareness methods-users seem to have of this dark side. Increasingly assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results. 
Additionally, little consideration is given to whether using a PCM is really appropriate for the question at hand \citep{losos2011seeing}. 
To paraphrase \citet{blomberg2012independent}, the enthusiastic application of PCMs has greatly outpaced our theoretical understanding of the methods and their relations to the rest of statistical theory.

@@

One major issue is that two groups of people are involved with PCMs. 
At one extreme, methods-developers (wookies) are involved with developing new methods and implementing them. They are aware of the limitations of their methods, and the assumptions that underly them. 
% Bit of a stretch but we can pretend right? :)
At the other extreme, naive methods-users (ewoks) have no knowledge of these complexities, they just want to use PCMs on their data. 
They may be graduate students or more established scientists who have been asked to apply a phylogenetic correction by a reviewer. 
Of course the majority of PCM users fall somewhere between these two extremes on the ``wookie-ewok'' spectrum; many methods-developers are developing methods with a specific empirical question in mind, while there are many advanced methods-users who have detailed knowledge of the limitations of the methods they use. 
Although methods-developers (and advanced methods-users) are aware of the problems with PCMs, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers, but who should take responsibility for such actions? And why has this gap developed in the first place? Here we explore these issues and make suggestions for improvements.\\ %needs some rewording

>>>>

A long time ago in a galaxy far, far away - also known as the 1980s - phylogenetic comparative methods (PCMs) were developed to deal with the statistical non-independence of species in comparative analyses (e.g. \citealp{felsenstein1985phylogenies,grafen1989phylogenetic}). 
Since then PCMs have been extended to investigate evolutionary pattern and process, and include methods for investigating drivers of diversification, the tempo and mode of trait evolution, and models of speciation and extinction (see reviews in \citealp{o2012evolutionary, pennell2013integrative}). 
PCMs have also become extremely popular over recent years; the number of papers containing the phrase `phylogenetic comparative' has increased exponentially since the 1980s (Figure \ref{PCMCitations}). 
With new methods being published almost weekly, there has never been a better time to be a comparative biologist.

\begin{figure}[h]
\centering
\includegraphics[width = 10cm]{PCMCitations.png}
\caption{PCM papers through time}
\label{PCMCitations}
\end{figure}

Unfortunately, PCMs also have a ``dark side''; they make various assumptions and suffer from biases in exactly the same way as any other statistical method \citep{freckleton2009seven,boettiger2012your}. 
The more popular these methods become, the less awareness methods-users seem to have of this dark side. Increasingly assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results. 
Additionally, little consideration is given to whether using a PCM is really appropriate for the question at hand \citep{losos2011seeing}. 
To paraphrase \citet{blomberg2012independent}, the enthusiastic application of PCMs has greatly outpaced our theoretical understanding of the methods and their relations to the rest of statistical theory.

@@

One major issue is that two groups of people are involved with PCMs. 
At one extreme, methods-developers (``wookies'') are involved with developing new methods and implementing them. They are aware of the limitations of their methods, and the assumptions that underly them. 
% Bit of a stretch but we can pretend right? :)
At the other extreme, naive methods-users (``ewoks'') have no knowledge of these complexities, they just want to use PCMs on their data. 
They may be graduate students or more established scientists who have been asked to apply a phylogenetic correction by a reviewer. 
Of course the majority of PCM users fall somewhere between these two extremes on the ``wookie-ewok'' spectrum; many methods-developers are developing methods with a specific empirical question in mind, while there are many advanced methods-users who have detailed knowledge of the limitations of the methods they use. 
Although methods-developers (and advanced methods-users) are aware of the problems with PCMs, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers, but who should take responsibility for such actions? And why has this gap developed in the first place? Here we explore these issues and make suggestions for improvements.\\ %needs some rewording
